---
title: "M06-Reflection Essay-Advanced Data Wrangling"
author: "Danielle Meza"
date: today
format:
  html:
    toc: true
    toc-depth: 4
    toc-expand: 2
    number-sections: true
    code-link: true
    code-fold: true
    code-tools: true
    code-overflow: wrap
    theme: cosmo
    embed-resources: true
editor: visual
execute:
  freeze: auto
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(knitr)
```

# Revealjs Slides

## My Impression

My first impression of Revealjs was how *professional and minimal* the slides look. Compared with the busy templates I usually see in PowerPoint, the clean layouts let the content stand out, and the soft animated transitions between slides make a presentation feel polished without being distracting. What impressed me most is that all of this is controlled by a few lines of text. The look and feel, including the style, transitions, and layout, comes from settings in the YAML header under `revealjs`, so I can change the whole presentation by editing one option instead of reformatting slide by slide.

## Capabilities

From the example presentations, Revealjs slides can do much more than static slides. Because they are built from plain text and rendered as a web page, they can:

-   mix **narrative, live R code, tables, and charts** in one source file, so every result is reproducible;
-   use **themes, transitions, incremental lists, and fragments** to control how content appears;
-   arrange content in **columns** and add **speaker notes**, **footers**, and **logos**;
-   be published to the web, so anyone can open the slides with a single link.

## Revealjs vs. PowerPoint

::: {.panel-tabset}
### Strengths

1.  **Reproducibility:** charts and tables are generated from code, so they update automatically when the data change.
2.  **Consistency:** one theme and one YAML header control the look of every slide.
3.  **Version control:** a text file works well with Git, so changes are easy to track.
4.  **Sharing:** a link opens the deck in any browser, with no software to install.

### Weaknesses

1.  **Learning curve:** there is no drag-and-drop editor, so layout takes practice.
2.  **Fine-tuning visuals:** custom animations and exact positioning are easier in PowerPoint.
3.  **Collaboration:** teammates who do not use Quarto cannot easily co-edit a deck.
4.  **Required formats:** many organizations still expect a `.pptx` file.
:::

Overall, I see Revealjs as the better choice when a presentation includes analysis that must stay reproducible, and PowerPoint as the better choice when quick visual editing or collaboration with non-technical teammates matters most.[^1]

[^1]: A practical tip: I started from the professor's example files and changed one setting at a time, which made it easy to see what each option did.

# My Presentation

## Link

[**View my published Revealjs presentation**](https://01a10a37-d710-1a1c-39c9-fa88411ef662.share.connect.posit.cloud/){target="_blank"}

## Topic

The presentation is a proposal talk for Analytics Objective 1 (AO1) of my Culminating Experience Project with the CPP Farm Store. It follows the storytelling structure from the assignment: a problem statement, the analytics objective and why it matters, a literature review on omnichannel integration and inventory visibility, and the planned methods. AO1 asks whether the Farm Store's new online storefront, integrated with its Clover point-of-sale system, performs well and where friction points limit online sales.

The presentation includes a reproducible table and a chart, both built from R code that runs when the slides render. Because Clover data access has not been confirmed yet, they use [simulated data]{.mark} that I labeled "Illustrative" on the slides, so they show the planned analysis without being presented as real Farm Store results. The slides also compute the sample size with inline R code instead of typing it by hand.

## What I Learned

Building this presentation showed me how much control the YAML header gives me. Options like `transition`, `transition-speed`, `theme`, and `footer` set the behavior and style of the whole deck, and I could still override a transition on a single slide (I used fade and zoom on a few) when I wanted variety. I also learned that a presentation can contain real, reproducible analysis. The table and chart come from R code that runs when the slides render, and the sample size in my narrative is calculated with inline code.

The biggest takeaway is how useful this will be going forward. Publishing the slides to the web also made sharing easy, because anyone can open one link. I am excited to use Revealjs for future projects and especially for my year-long CEP, where I can reuse the same theme and update the charts automatically once the Farm Store data become available.

# Advanced Wrangling

I chose one function from each of three categories in the lecture: **Relational Data & Joins** (`left_join()`), **Rankings** (`min_rank()`), and **Conditional Statements** (`case_when()`). To keep the examples reproducible, I built small Farm Store tables directly in R.

## Example Data

::: {.panel-tabset}
### `orders`

```{r}
#| label: tbl-orders
#| tbl-cap: "Sample online orders"
orders <- tribble(
  ~order_id, ~product_id, ~qty,
  1, "P1", 2,
  2, "P2", 1,
  3, "P1", 3,
  4, "P4", 1
)
kable(orders)
```

### `products`

```{r}
#| label: tbl-products
#| tbl-cap: "Sample product catalog"
products <- tribble(
  ~product_id, ~product_name, ~category,
  "P1", "Farm Eggs (12 ct)", "Eggs",
  "P2", "Strawberries (1 lb)", "Produce",
  "P3", "Succulent Plant (4 in)", "Plants"
)
kable(products)
```
:::

The first example uses @tbl-orders and @tbl-products.

## Joining: `left_join()` {#sec-join}

**When to use it:** use `left_join()` when one table is the main table (here, `orders`) and you want to add matching information from a second table (here, `products`) without losing any rows from the main table. This is a core relational-data skill, because details such as orders and products are usually stored in separate tables linked by a shared key.

```{r}
#| label: tbl-left-join
#| tbl-cap: "Orders joined to product details with left_join()"
orders_joined <- orders |>
  left_join(products, by = "product_id")
kable(orders_joined)
```

**Reading the output:** @tbl-left-join keeps all four orders. Order 4 uses product `P4`, which is not in the catalog, so its `product_name` and `category` are `NA`. An `inner_join()` would have silently dropped that order. That is why I would choose `left_join()` for the Farm Store: it exposes the data-quality problem instead of hiding it.

::: {.callout-tip}
## Tip
Joining on a shared key such as `product_id` is the R version of the `JOIN ... ON` pattern I practiced in SQL and BigQuery.
:::

## Rankings: `min_rank()` {#sec-rank}

**When to use it:** use `min_rank()` when you need to rank rows from best to worst, such as top-selling categories, and you want tied values to share the same rank. It is one of the ranking functions from the lecture, alongside `row_number()` and `dense_rank()`.

```{r}
#| label: tbl-rank
#| tbl-cap: "Category sales ranked with min_rank()"
category_sales <- tribble(
  ~category, ~total_sales,
  "Eggs",     405,
  "Produce",  305,
  "Honey",    305,
  "Plants",   175
)

category_ranked <- category_sales |>
  mutate(sales_rank = min_rank(desc(total_sales)))
kable(category_ranked)
```

**Reading the output:** `desc()` makes the highest sales rank first. Produce and Honey tie at 305, so both receive rank 2, and the next category (Plants) receives rank 4, because `min_rank()` leaves a gap after a tie. `dense_rank()` would give Plants rank 3 instead. I would use `min_rank()` when the gap is meaningful, for example "two categories outrank Plants" versus "Plants is third-best."

## Conditions: `case_when()` {#sec-case}

**When to use it:** use `case_when()` to create a new variable from several if/else rules at once. It is easier to read than a chain of nested `if_else()` calls, and the rules are checked in order, so the first one that is true wins.

```{r}
#| label: tbl-case
#| tbl-cap: "Sales tiers created with case_when()"
category_tiers <- category_ranked |>
  mutate(
    tier = case_when(
      total_sales >= 350 ~ "Top seller",
      total_sales >= 250 ~ "Steady",
      TRUE               ~ "Needs attention"
    )
  )
kable(category_tiers)
```

**Reading the output:** Eggs (405) lands in "Top seller," Produce and Honey (305) in "Steady," and Plants (175) falls through to the final `TRUE` rule, "Needs attention." The last `TRUE ~` line is the catch-all for anything that matched none of the earlier rules. A grouping like this could help a store manager see which categories deserve promotion or a closer look.

## Summary

| Function | Lecture category | Use it when |
|:---------------|:---------------|:-------------------------------------------|
| `left_join()` | Relational data & joins | You need to add details from another table and keep every row of the main one |
| `min_rank()` | Rankings | You need to rank rows and let ties share a rank |
| `case_when()` | Conditional statements | You need a new variable based on several rules |

: Summary of the three functions {#tbl-summary}

@tbl-summary shows how the three functions fit a typical analysis workflow: join the data (@sec-join), rank the results (@sec-rank), and label them with business rules (@sec-case). For my capstone, joining supports combining Clover and online store data, ranking supports finding top and bottom SKUs, and `case_when()` supports tiering categories for the dashboards.

::: {.callout-important}
## Takeaway
Each of these functions turns raw tables into something a decision-maker can act on: a complete combined dataset, a ranked list, and a labeled priority.
:::