---
title: "R Markdown: Reproducible Report in Data Science"
author: "Danielle Meza"
date: "`r format(Sys.time())`"
output:
  html_document:
    code_folding: show
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: false
    highlight: espresso
    theme: journal
    number_sections: yes
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Markdown

R Markdown is a format for writing reproducible, dynamic reports with R.
Use it to embed R code and results into slideshows, PDFs, HTML
documents, Word files, and more.

# Structure

1.  YAML Header
2.  Text Area
3.  Code chunk

# Workflow

1.  **Open** – Open a file that uses the .Rmd extension.
2.  **Write** – Write content with the easy-to-use R Markdown syntax.
3.  **Embed** – Embed R code that creates output to include in the
    report.
4.  **Render** – Replace R code with its output and transform the report
    into a slideshow, PDF, HTML, or MS Word file (done by clicking
    **Knit**).

# Text Area Formatting Tools

Plain text just types normally. End a line with two trailing spaces to
force a new paragraph/line break.

*italics* and *italics*

**bold** and **bold**

superscript^2^

~~strikethrough~~

Hyperlink:
[CCIDM](https://www.cpp.edu/cba/customer-insights-lab/index.shtml)

# Header 1

## Header 2

### Header 3

#### Header 4

##### Header 5

###### Header 6

endash: -- emdash: --- inline equation: $A = \pi*r^2$ slide break:

------------------------------------------------------------------------

> Block quote

1.  Ordered list 1
2.  Ordered list 2
    -   sub-item 1
    -   sub-item 2

| Table Header | Second Header |
|:------------:|--------------:|
|  Table Cell  |        Cell 2 |
|    Cell 3    |        Cell 4 |

# YAML Header Explained

> A YAML header is a set of key: value pairs at the start of your file.
> Begin and end the header with a line of three dashes (`---`). (From
> the Rmd cheatsheet:
> <https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf>)

My YAML header above is different from the plain default RStudio
generates automatically — that's because I customized the `output`
options, and those settings control how the knitted HTML file will look.

-   **date**: `` `r format(Sys.time())` ``
    -   This prints the current date/time every time I knit the file,
        instead of a hardcoded date.
    -   Can be customized several ways:
        -   date-time: `` `r Sys.time()` ``
        -   Day, Month, date, Year:
            `` `r format(Sys.time(), "%A, %B, %d, %Y")` ``
        -   m/d/Y format: `` `r format(Sys.time(), "%m/%d/%Y")` ``

------------------------------------------------------------------------

-   **code_folding**: show
    -   Lets readers collapse/expand code chunks in the knitted HTML,
        but shows the code by default.
-   **toc**: yes
    -   Turns on the table of contents. It gets hyperlinks in the HTML
        output so readers can jump to sections quickly.
-   **toc_depth**: 4
    -   Table of contents will show up to 4 levels of headings deep.
-   **toc_float**: (nested, collapsed: false)
    -   Moves the TOC from the top of the document to a floating sidebar
        that stays visible as you scroll.
    -   By default the floating TOC starts collapsed (not showing all
        headings) — setting `collapsed: false` under it keeps it
        expanded.
-   **highlight**: espresso
    -   Controls the syntax-highlight style of code chunks — espresso
        gives a black background for the code area.
-   **theme**: journal
    -   Overall visual theme of the knitted document. Lots of built-in
        options to try (cosmo, flatly, journal, united, etc.).
-   **number_sections**: yes
    -   Auto-numbers your headings, which is why I don't need to
        manually number sections above (e.g. "1 R Markdown", "2
        Structure", etc.)
    -   Try swapping these values and re-knitting to see what changes.

# Embed Codes

## Code chunk

### Sample

Insert a new code chunk with **Cmd + Option + I** (Mac)

```{r penguins, message=FALSE, warning=FALSE}
# use ctl + Enter simultaneous to execute a line
# Use # to comment out a code

# install.packages("tidyverse")
library(tidyverse)

# install.packages("palmerpenguins")
library(palmerpenguins)
penguins

dim(penguins)

# install.packages("skimr")
skimr::skim(penguins)

# Sex distribution
penguins %>%
  count(sex) %>%
  ggplot(aes(sex, n)) +
  geom_col()

# Number of male
sum(penguins$sex == "male", na.rm = TRUE)

# Total observation
nrow(penguins)

# Percent of Males
m_per <- round((sum(penguins$sex == "male", na.rm = TRUE) / nrow(penguins)) * 100, digits = 1)

# Percent of Females
f_per <- round((sum(penguins$sex == "female", na.rm = TRUE) / nrow(penguins)) * 100, digits = 1)
```

### Code chunk options

Chunk options (like `message=FALSE, warning=FALSE` in the
`{r penguins, ...}` header above) control what shows up in the knitted
output for that specific chunk — e.g. suppressing package-load
messages/warnings so they don't clutter the report. For the full list of
options, see <https://yihui.org/knitr/options/>

## Inline coding

Inline code lets you drop a single R value directly into a sentence of
regular text, instead of only showing results inside a chunk. Syntax is
a single backtick, the letter `r`, your code, then a closing backtick.

-   Sample characteristics: There are `r nrow(penguins)` penguins in the
    data, `r m_per`% of which are males and `r f_per`% of which are
    females.

(Notice this sentence will always stay accurate even if the underlying
data changes — that's the "reproducible" part of reproducible reporting:
re-knit and the numbers update automatically instead of me manually
retyping them.)

# Navigation

-   Outline (right side) — jumps between headings in your source file
-   Status bar (bottom) — shows your current cursor location/section

# Expectations for Organizing Your Assignment Report with an Rmd File

## DS-L01-2 Rmd Principle Assignment

(The instructor's own version of this template says "M01-1 Principle
Assignment" here, since he reuses this file across different courses —
swap in your actual assignment name so the heading matches this
assignment.)

# References

-   R Markdown cheatsheet:
    <https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf>
-   knitr chunk options: <https://yihui.org/knitr/options/>
